---
title: "Swedish Birds"
author: "Markus Skyttner"
date: "January 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Swedish Birds Data

Let us have a look at two packages provide some dataset with long term data about Swedish Birds:

- swedishbirdtrends
- swedishbirdrecoveries

Let's see how we can use these packages..

## Exercises on "swedishbirdtrends"

- Load the dataset in the package "swedishbirdtrends"
	- How do you load the package?
	- Why do you think that this packages are not available from ROpenSci or CRAN?
	- Where can you get these packages?
		- swedishbirdtrends at https://github.com/mskyttner/swedishbirdtrends
	
- There is some data in the package, summarize it briefly
	- Is it large?
	- What number of rows and columns?
	- Is it up-to-date?
	- Where can you find information about how to use the package?
	- Where does the bundled datasets come from?
	- Where can you find the source code for this package?
	- Can you say anything about where some output form this package is available? (hint: https://github.com/mskyttner/swedishbirdtrends/issues/1)
	- How do you think this package is intended to be used?
	
```{r}

library(swedishbirdtrends)

birdtrends
birdtotals

nrow(birdtrends)
ncol(birdtrends)
names(birdtrends)

```

## Exercises on "swedishbirdrecoveries"

- Where can you get this packages?
	- swedishbirdrecoveries at https://github.com/mskyttner/swedishbirdrecoveries
	- Why do you think that this packages are not available from ROpenSci or CRAN?

- There is some data in the package, summarize it briefly
	- How many datasets? Why?
	- Give some stats on the size of the datasets?
	- What number of rows and columns?
	- Size on disk?
	- Is this dataset up-to-date? Why?
	- Where can you find information about how to use the package?
	- Where does the bundled datasets come from?
	- Can you say anything about where some output form this package is available?
	- How do you think this package is intended to be used?
	
## Web UI	for swedish bird recoveries
	
There is a web app at https://birdringing.dina-web.net 

- Can you say discover something about the dataset based on it?
- Can you suggest some changes and improvements?
- Where could you file such change requests or bugs or issues that you discover?
- Would any rOpenSci packages be relevant for usage with this dataset?
- Could any other package mentioned here or otherwise be relevant to use?